Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia

### 2024-08-19 manuscript/manuscript_word.Rmd
finished another round of proofreading.

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract:
|
  ```{=latex}
  The appearance of ‘Large Cutting Tools’ in the early Acheulean is widely regarded as the first evidence for the imposition of intended forms on artifacts, with major implications for hominin cognitive and cultural capacities. However, the nature and extent of explicit design documented by these forms remains open to debate. To address this issue, we analyzed the complete collection of early Acheulean (ca. 1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared these with all of the flaked pieces from two published Oldowan (>2.5 Ma) sites at Gona. By comparing shape variation to measures of flaking intensity and coverage, we sought to identify technological patterns indicative of intent. Results provide little support for the presence of discrete tool types or imposed morphological norms in our sample. Instead, we observe systematic patterns of raw material selection and core surface modification aimed at the production and maintenance of useful cutting edges on relatively large supports (cobbles and large flake blanks). This is consistent with prior characterizations of Acheulean tool form as arising from functional and ergonomic design imperatives for large hand-held cutting tools. Although the generalizability of these results to other sites remains to be seen, we propose that distinctive early Acheulean artifact forms may represent secondary accommodations to the primary goal of increasing tool size to meet the novel demands (e.g., extended use-life, enhanced transportability, utility for heavy-duty cutting) of a more general shift in hominin behavioral ecology at this time. Our results thus provide support for the presence of Acheulean design, not as the concrete realization of a ‘holographic model’ in the mind of the maker, but rather in the broader sense of deliberate technological choices made in view of behavioral goals and material constraints.

  \textbf{Keywords:} Lithic technology; Early Stone Age; Handaxe; Large Cutting Tool; Pick; Shaping
  ```
date: "`r Sys.Date()`"
fontsize: 11pt
output: bookdown::pdf_document2
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: jhe.csl
---

>>>>

---
title: "Imposed form in the Early Acheulean? Evidence from Gona, Afar, Ethiopia"
author:
- Dietrich Stout^[Department of Anthropology, Emory University, Atlanta, GA, USA;  dwstout@emory.edu]
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
- Antoine Muller^[Institute of Archaeology, Mount Scopus, The Hebrew University of Jerusalem, Jerusalem, Israel; antoine.muller@mail.huji.ac.il]
- Michael J. Rogers^[Department of Anthropology, Southern Connecticut State University, New Haven, CT, USA; rogersm1@southernct.edu]
- Sileshi Semaw^[Centro Nacional de Investigación sobre la Evolución Humana (CENIEH), Burgos, Spain; sileshi.semaw@cenieh.es]
abstract: The appearance of ‘Large Cutting Tools’ in the early Acheulean is widely regarded as the first evidence for the imposition of intended forms on artifacts, with major implications for hominin cognitive and cultural capacities. However, the nature and extent of explicit design documented by these forms remains open to debate. To address this issue, we analyzed the complete collection of early Acheulean (ca. 1.7–1.2 Ma) flaked pieces from four sites (BSN17, DAN5, OGS12, and OGS5) in the Gona Project Area and compared these with all of the flaked pieces from two published Oldowan (>2.5 Ma) sites at Gona. By comparing shape variation to measures of flaking intensity and coverage, we sought to identify technological patterns indicative of intent. Results provide little support for the presence of discrete tool types or imposed morphological norms in our sample. Instead, we observe systematic patterns of raw material selection and core surface modification aimed at the production and maintenance of useful cutting edges on relatively large supports (cobbles and large flake blanks). This is consistent with prior characterizations of Acheulean tool form as arising from functional and ergonomic design imperatives for large hand-held cutting tools. Although the generalizability of these results to other sites remains to be seen, we propose that distinctive early Acheulean artifact forms may represent secondary accommodations to the primary goal of increasing tool size to meet the novel demands (e.g., extended use-life, enhanced transportability, utility for heavy-duty cutting) of a more general shift in hominin behavioral ecology at this time. Our results thus provide support for the presence of Acheulean design, not as the concrete realization of a ‘holographic model’ in the mind of the maker, but rather in the broader sense of deliberate technological choices made in view of behavioral goals and material constraints.
  \textbf{Keywords:} Lithic technology; Early Stone Age; Handaxe; Large Cutting Tool; Pick; Shaping
date: "`r Sys.Date()`"
fontsize: 11pt
output: word_document
header-includes:
   - \usepackage{textcomp}
   - \usepackage{utopia}
   - \usepackage{lineno}
   - \usepackage[utf8]{inputenc}
   - \linenumbers
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage{caption}
   - \captionsetup[figure]{font=scriptsize}
   - \captionsetup[table]{font=scriptsize}
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: jhe.csl
---
